\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics

### 2018-12-22 Documents/Euler/Astronum_2018/ms_endeve.tex
Work on response

The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.  
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\ is based on the DG method, while we foresee using a combination of spectral and continuous finite element methods for solving elliptic equations; e.g., for Newtonian gravity or relativistic gravity employing the conformal flatness approximation \cite{wilson_etal_1996}.  
With these approaches, we employ high-order discretization techniques based on a common mathematical framework in all the major model components.  
In this paper, we provide an initial description and encouraging demonstration of the DG method implemented in \thornado\ to solve the Euler equations.  
We note that others (e.g., \cite{radiceRezzolla_2011,schaal_etal_2015}) have also investigated the DG method in the context of astrophysical hydrodynamics and reported encouraging results.  
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.

@@

Note that the interpolation points $S_{N}$ and the quadrature points $\hat{S}_{M}$ generally do not coincide.  
However, for the sake of computational efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a spectral-type nodal collocation DG approximation \cite{bassi_etal_2013}.  
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG}, letting $v(x)=\ell_{k}(x)$, and using the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\pd{}{t}\vect{U}_{k}\,\dx
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative term.  
For a general metric, the integral in Eq.~\eqref{eq:timeDerivativeTerm} is approximate since we use the Legendre-Gauss quadrature rule with the nodal points given by the expansion in Eq.~\eqref{eq:conservedNodalExpansion}.  
However, since $\ell_{i}(\eta_{k})=\delta_{ik}$, this leads to a diagonal mass matrix, and simplifies the implementation.  
Similarly, we obtain
\begin{align}
  \int_{K}\vect{S}(\vect{U}_{h})\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\vect{S}(\vect{U}_{k})\,\dx.
  \label{eq:sourceTerm}
\end{align}
for the source term.  
Finally, the volume term (last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}) becomes
\begin{equation}
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{v}{x}\,dV
  \approx \sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\end{equation}
Using Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in Eq.~\eqref{eq:semidiscreteDG} results in the semi-discrete form
\begin{align}
  \deriv{\vect{U}_{k}}{t}
  =-\f{1}{w_{k}\sqrt{\gamma}_{k}\dx}
  \Big\{
  \Big[\,
    \sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xH}
    -\sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
  -\sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\} + \vect{S}(\vect{U}_{k}).
  \label{eq:semidiscreteDiscretized}
\end{align}
Eq.~\eqref{eq:semidiscreteDiscretized} comprises a system of ordinary differential equations (ODEs), which are integrated in time with an ODE solver.  
In Section~\ref{sec:numerical} we use the optimal third-order strong stability-preserving Runge-Kutta (SSP-RK3) method from \cite{shuOsher_1988}.  
We assume that the source term $\vect{S}(\vect{U}_{k})$ is non-stiff so that the equations can integrated efficiently with explicit methods.  
In future applications of the DG method to model CCSNe we plan to use a combination of implicit and explicit methods (e.g., IMEX methods \cite{ascher_etal_1997}) to stably integrate stiff sources due to neutrino-matter interactions.

Limiting of the polynomial representation $\vect{U}_{h}$, to reduce oscillations and prevent non-physical states, is a critical step in the DG algorithm.  
We use the TVD-type slope limiter based on limiting characteristic variables, discussed in \cite{cockburnShu_1998}.  
To prevent excessive limiting (e.g., at smooth extrema), we use the troubled-cell indicator (TCI) of \cite{fuShu_2017}:
\begin{equation}
  I_{\bK}(G) = \f{\sum_{j}|G_{\bK}-G_{\bK}^{(j)}|}{\max_{j}|G_{\bK^{(j)}}^{(j)}|},
  \label{eq:indicator}
\end{equation}
where $G\in\vect{G}\subseteq\vect{U}$.  
(Here, $\vect{G}$ consists of the first and last element of $\vect{U}$.)  
In Eq.~\eqref{eq:indicator}, the sum in the numerator and the $\max$ in the denominator extend over the neighboring elements sharing a face with the target element $\bK$.  
$G_{\bK}$ is the cell average in $\bK$, $G_{\bK}^{(j)}$ is the cell average computed by extrapolating the polynomial representation from the neighbor element $\bK^{(j)}$ into $\bK$, and $G_{\bK^{(j)}}^{(j)}$ is the cell average native to $\bK^{(j)}$.  
An element is flagged for limiting if $I_{\bK}(G)>C_{\TCI}$ for any $G\in\vect{G}$.  
To prevent negative mass density and pressure (and superluminal velocity for the relativistic Euler equations), we follow the approaches in \cite{zhangShu_2010,wuTang_2015,qin_etal_2016}.

To summarize the RK-DG algorithm we let $\bar{\vect{U}}$ denote the global solution vector containing all the unknowns in $D$ and $\bar{\vect{F}}$ denote the corresponding DG discretization from the right-hand side of Eq.~\eqref{eq:semidiscreteDiscretized}, so that the global ODE system can be written as the $d_{t}\bar{\vect{U}}=\bar{\vect{F}}(\bar{\vect{U}})$.  
We denote the TVD limiter with $\Lambda^{\TVD}\{\}$ and the `positivty' limiter preventing unphysical states with $ \Lambda^{\Pos}\{\}$.  
(E.g., application of the TVD limiter is $\bar{\vect{U}} := \Lambda^{\TVD}\{\bar{\vect{U}}\}$.)  
The general $s$-stage Runge-Kutta time stepping algorithm, which includes application of limiters, can then be written as \cite{cockburnShu_2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
    \bar{\vect{U}}^{(i)} 
    = \Lambda^{\Pos}\Big\{\,\Lambda^{\TVD}\Big\{\,\sum_{j=0}^{i-1}\Big(\,\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\,\Big)\,\Big\}\,\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.  
\end{itemize}
The coefficients $\alpha_{ij}$ and $\beta_{ij}$ for SSP-RK3 are given in Table~2.1 in \cite{cockburnShu_2001}.  
In Eq.~\eqref{eq:rkStages}, for each stage, the TVD limiter is applied first (for elements where $I_{\bK}>C_{\TCI}$), followed by the positivity limiter.  
The positivity limiter

$\Delta t=\min_{D}(\Delta t_{\TVD},\Delta t_{\Pos})$
\begin{equation}
  \Delta t_{\TVD} = \f{1}{d}\,\f{1}{(2k+1)}\min_{i}\big(\Delta x^{i}/|\lambda^{i}|\big)\quad\text{and}\quad
  \Delta t_{\Pos} = \f{\hat{w}_{N}}{d}\min_{i}\big(\Delta x^{i}/|\lambda^{i}|\big)
\end{equation}
For $d=k=2$, the effective Courant-Friedrichs-Lewy (CFL) factor is $0.1$ for $\Delta t_{\TVD}$, while it is $1/12$ for $\Delta t_{\Pos}$.

@@

In this section we show preliminary numerical results obtained with the DG method implemented in \thornado\ to solve the non-relativistic and special relativistic Euler equations.  
We use the ideal gas equation of state where the pressure $p$ is related to the internal energy density $e$ by $p=(\Gamma-1)e$, and where $\Gamma$ is the (constant) adiabatic index.  
Let the $d$-dimensional computational domain be $D=\times_{i=1}^{d}[\xMin^{i},\xMax^{i}]$, where $\xMin^{i}$ and $\xMax^{i}$ denote the coordinates of the inner and outer boundary in the $i$th coordinate direction, respectively.  
To limit the scope of this initial study, we use polynomials of degree $k=2$ (except in Section~\ref{sec:isentropicVortex}, where we also use $k=1$ and $k=3$) combined with SSP-RK3 time stepping, and run all the tests with a Courant-Friedrichs-Lewy (CFL) factor of $C=0.1$.  
We use the HLL Riemann solver when solving the non-relativistic Euler equations and the HLLC Riemann solver when solving the special relativistic Euler equations.  
The purpose of the tests is to gauge the performance of the DG implementation on a set of benchmarks as an initial measure of its suitability for future CCSN simulations.

@@

The first test is the classic Riemann problem due to Sod \cite{sod_1978}, computed using Cartesian coordinates.  
The one-dimensional computational domain is $D=[0,1]$, and a discontinuity is initially located at $x^{1}=0.5$, separating the left and right states
\begin{equation*}
  \vect{U}_{\leftState}=(1.0,\vect{0.0},2.5)^{\trans}\quad\text{and}\quad
  \vect{U}_{\rightState}=(0.125,\vect{0.0},0.25)^{\trans},
\end{equation*}
where the adiabatic index is $\Gamma=1.4$.  
The test is run with $100$ elements until $t=0.2$, using $C_{\TCI}=0.03$.  
Results are plotted in the left panels of Figure~\ref{fig:SodSedov}.  
The DG method captures the main features of the solution well without introducing noticeable oscillations near the discontinuities (upper left panel).  
In the lower left panel we plot locations in the $xt$-plane of elements flagged for limiting.  
Limiting is mainly triggered by the contact discontinuity and the shock.  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sod_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sedov_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sod_TCI_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sedov_TCI_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:SodSedov}{\it Upper left panel:} numerical solution of Sod's problem using $100$ elements (plusses) compared with the exact solution (solid lines; from \cite{toro_1999}).  {\it Upper right panel:} numerical solution of the Sedov-Taylor blast wave using $256$ elements (plusses) against the exact solution (solid lines).  In the lower panels we plot locations in the $xt$-plane of elements flagged for limiting ($I_{\bK}>C_{\TCI}$) by the Sod (left) and Sedov (right) problems.}
\end{figure}

@@

\begin{wrapfigure}{l}{2.825in}
  \includegraphics[width=18pc]{./Figures/IsentropicVortex_Astronum_2018}
  \caption{\label{fig:IsentropicVortex}Convergence results for the Isentropic Vortex:  $L^{\infty}$ error norm versus number of elements in the $x^{1}$-dimension.}
\end{wrapfigure}

In Figure~\ref{fig:IsentropicVortex} we plot the density error measured in the infinity norm $L^{\infty}=\max_{D}|\rho(\vect{x},t_{f})-\rho_{0}(\vect{x})|$ versus number of cells the $x^{1}$-dimension for various configurations.  
We ran with spatial resolutions of $25\times25$, $50\times50$, $100\times100$, and $200\times200$.  
In the first set of runs, we ran without any limiting and varied the polynomial degree $k$: $k=1$ (solid red), $k=2$ (solid black), and $k=3$ (solid green).  
We observe the expected order of accuracy for $k=1$ and $k=2$ (second- and third-order accuracy, respectively; cf. dotted reference lines).  
For $k=3$, the order of accuracy is close to fourth-order for the lower resolutions, but tends to third-order as the resolution increases, presumably due to the use of third-order time stepping.  
In the second set of runs, we kept $k=2$ and ran with limiting turned on.  
We ran one set with $C_{\TCI}=0.03$ (dashed magenta) and one set with $C_{\TCI}=0.1$ (dashed cyan).  
When $C_{\TCI}=0.03$, the limiter is activated for the lowest resolution run and significantly reduces the accuracy of the method.  
For all the other runs, the limiter is not activated, and the results are identical to the runs with the limiter turned off.

@@

\begin{wrapfigure}{l}{2.75in}
  \centering
  \includegraphics[width=16pc]{./Figures/KelvinHelmholtz_TCI_24_Astronum_2018}
  \caption{\label{fig:KelvinHelmholtz_TCI}Distribution of elements flagged for limiting in the KH test at $t=2.4$.}
\end{wrapfigure}

@@

\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Implosion_01_HighTCI_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Implosion_02_HighTCI_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:Implosion}Plots for the Liska-Wendroff implosion problem at $t=0.045$ (left) and $t=2.5$ (right), intended to match Figures~4.10 and 4.11 in \cite{liskaWendroff_2003}.  The color map shows the pressure distribution, black contours show the mass density, while arrows indicate the velocity.}
\end{figure}

>>>>

The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.  
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\ is based on the DG method, while we foresee using a combination of spectral and continuous finite element methods for solving elliptic equations; e.g., for Newtonian gravity or relativistic gravity employing the conformal flatness approximation \cite{wilson_etal_1996}.  
With these approaches, we employ high-order discretization techniques based on a common mathematical framework in all the major model components.  
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.  
However, in this paper, we provide an initial description and encouraging demonstration of the DG method implemented in \thornado\ to solve the Euler equations.  
We focus on basic tests in one and two spatial dimensions and focus, to some extent, on the implementation of limiters in \thornado.  
We note that others (e.g., \cite{radiceRezzolla_2011,schaal_etal_2015,zanotti_etal_2015,dumbser_etal_2018}) have also investigated the DG method in the context of astrophysical applications and reported encouraging results.  
The DG limiting techniques implemented in \thornado\ are similar to those used by \cite{schaal_etal_2015}, but is supplemented with a troubled-cell indicator \cite{fuShu_2017} to prevent excessive limiting (e.g., around smooth extrema), and differs from more recent approaches in the literature (e.g., \cite{huerta_etal_2012,dumbser_etal_2014,zanotti_etal_2015,sonntagMunz_2017}), which combine the sub-cell resolution of the DG method with high-order finite-volume techniques to improve the performance of the DG limiting procedure.  
While the results with current limiters are encouraging, these latter methods appear as attractive options for future implementation in \thornado.

@@

Note that the interpolation points $S_{N}$ and the quadrature points $\hat{S}_{M}$ generally do not coincide.  
However, for the sake of computational efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a spectral-type nodal collocation DG approximation \cite{bassi_etal_2013}.  
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG}, letting $v(x)=\ell_{k}(x)$, and using the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\pd{}{t}\vect{U}_{k}\,\dx
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative term.  
For a general metric, the integral in Eq.~\eqref{eq:timeDerivativeTerm} is approximate since we use the Legendre-Gauss quadrature rule with the nodal points given by the expansion in Eq.~\eqref{eq:conservedNodalExpansion}.  
However, since $\ell_{i}(\eta_{k})=\delta_{ik}$, this leads to a diagonal mass matrix, and simplifies the implementation.  
Similarly, we obtain
\begin{align}
  \int_{K}\vect{S}(\vect{U}_{h})\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\vect{S}(\vect{U}_{k})\,\dx.
  \label{eq:sourceTerm}
\end{align}
for the source term.  
Finally, the volume term (last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}) becomes
\begin{equation}
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{v}{x}\,dV
  \approx \sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\end{equation}
Using Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in Eq.~\eqref{eq:semidiscreteDG} results in the semi-discrete form
\begin{align}
  \deriv{\vect{U}_{k}}{t}
  =-\f{1}{w_{k}\sqrt{\gamma}_{k}\dx}
  \Big\{
  \Big[\,
    \sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xH}
    -\sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
  -\sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\} + \vect{S}(\vect{U}_{k}).
  \label{eq:semidiscreteDiscretized}
\end{align}
Eq.~\eqref{eq:semidiscreteDiscretized} comprises a system of ordinary differential equations (ODEs), which are integrated in time with an ODE solver.  
In Section~\ref{sec:numerical} we use the optimal third-order strong stability-preserving Runge-Kutta (SSP-RK3) method from \cite{shuOsher_1988}.  
We assume that the source term $\vect{S}(\vect{U}_{k})$ is non-stiff so that the equations can be integrated efficiently with explicit methods.  
In future applications of the DG method to model CCSNe we plan to use a combination of implicit and explicit methods (e.g., IMEX methods \cite{ascher_etal_1997}) to stably integrate stiff sources due to neutrino-matter interactions.

Limiting of the polynomial representation $\vect{U}_{h}$, to reduce oscillations and prevent unphysical states, is a critical step in the DG algorithm.  
We use the TVD-type slope limiter based on limiting characteristic variables, discussed in \cite{cockburnShu_1998}.  
To prevent excessive limiting (e.g., at smooth extrema), we use the troubled-cell indicator (TCI) of \cite{fuShu_2017}:
\begin{equation}
  I_{\bK}(G) = \f{\sum_{j}|G_{\bK}-G_{\bK}^{(j)}|}{\max_{j}|G_{\bK^{(j)}}^{(j)}|},
  \label{eq:indicator}
\end{equation}
where $G\in\vect{G}\subseteq\vect{U}$.  
(Here, $\vect{G}$ consists of the first and last element of $\vect{U}$.)  
In Eq.~\eqref{eq:indicator}, the sum in the numerator and the $\max$ in the denominator extend over the neighboring elements sharing a face with the target element $\bK$.  
$G_{\bK}$ is the cell average in $\bK$, $G_{\bK}^{(j)}$ is the cell average computed by extrapolating the polynomial representation from the neighbor element $\bK^{(j)}$ into $\bK$, and $G_{\bK^{(j)}}^{(j)}$ is the cell average native to $\bK^{(j)}$.  
An element is flagged for limiting if $I_{\bK}(G)>C_{\TCI}$ for any $G\in\vect{G}$.  
To prevent negative mass density and pressure (and superluminal velocity for the relativistic Euler equations), we follow the approaches in \cite{zhangShu_2010,wuTang_2015,qin_etal_2016}.

To summarize the full RK-DG algorithm we let $\bar{\vect{U}}$ denote the global solution vector containing all the unknowns in $D$ and $\bar{\vect{F}}$ denote the corresponding DG discretization from the right-hand side of Eq.~\eqref{eq:semidiscreteDiscretized}, so that the global ODE system can be written as the $d_{t}\bar{\vect{U}}=\bar{\vect{F}}(\bar{\vect{U}})$.  
The TVD limiter and the `positivity' limiter preventing unphysical states are denoted by $\Lambda^{\TVD}\{\}$ and $ \Lambda^{\Pos}\{\}$, respectively (e.g., application of the TVD limiter is $\bar{\vect{U}} := \Lambda^{\TVD}\{\bar{\vect{U}}\}$).  
The general $s$-stage Runge-Kutta time stepping algorithm, which includes application of limiters, can then be written as \cite{cockburnShu_2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
    \bar{\vect{U}}^{(i)} 
    = \Lambda^{\Pos}\Big\{\,\Lambda^{\TVD}\Big\{\,\sum_{j=0}^{i-1}\Big(\,\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\,\Big)\,\Big\}\,\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.  
\end{itemize}
The coefficients $\alpha_{ij}$ and $\beta_{ij}$ for SSP-RK3 are given in Table~2.1 in \cite{cockburnShu_2001}.  
In Eq.~\eqref{eq:rkStages}, for each stage, the TVD limiter is applied first (for elements where $I_{\bK}>C_{\TCI}$), followed by the positivity limiter.  
The algorithm is subject to two time step restrictions, the stability condition $\Delta t_{\TVD}$ and the positivity condition $\Delta t_{\Pos}$, and we take $\Delta t=\min_{D}(\Delta t_{\TVD},\Delta t_{\Pos})$.  
Let $(h/|\lambda|)=\min_{i\in\{1,\ldots,d\}}(\Delta x^{i}/|\lambda^{i}|)$, where $|\lambda^{i}|$ is the largest absolute eigenvalue of the flux Jacobian.  
Then, assuming Cartesian coordinates (e.g., \cite{cockburnShu_2001,zhangShu_2010}):
\begin{equation}
  \Delta t_{\TVD} \le \f{1}{d}\,\f{(h/|\lambda|)}{(2k+1)} \quad\text{and}\quad
  \Delta t_{\Pos} \le \f{\hat{w}_{1}(h/|\lambda|)}{d},
  \label{eq:timestepRestrictions}
\end{equation}
where $\hat{w}_{1}$ is the first quadrature weight of the Gauss-Lobatto quadrature that integrates the cell average of $\vect{U}_{h}$ exactly.  
For $d=k=2$, the effective Courant-Friedrichs-Lewy (CFL) factor is $0.1$ for $\Delta t_{\TVD}$, while it is $1/12$ for $\Delta t_{\Pos}$.  
The positivity limiter checks for unphysical states in all the quadrature points involved in the DG discretization.  
If any point in an element is flagged as unphysical, the polynomial representation $\vect{U}_{h}$ is damped towards its cell average $\vect{U}_{\bK}$, which is guaranteed to be physical by the $\Delta t_{\Pos}$ restriction.  
Note that $\Delta t_{\Pos}$ is a sufficient condition.  
In practice, the time step can often be set larger than $\Delta t_{\Pos}$.

@@

In this section we show preliminary numerical results obtained with the DG method implemented in \thornado\ to solve the non-relativistic and special relativistic Euler equations.  
We use the ideal gas equation of state where the pressure $p$ is related to the internal energy density $e$ by $p=(\Gamma-1)e$, and where $\Gamma$ is the (constant) adiabatic index.  
Let the $d$-dimensional computational domain be $D=\times_{i=1}^{d}[\xMin^{i},\xMax^{i}]$, where $\xMin^{i}$ and $\xMax^{i}$ denote the coordinates of the inner and outer boundary in the $i$th coordinate direction, respectively.  
To limit the scope of this initial study, we use polynomials of degree $k=2$ (except in Section~\ref{sec:isentropicVortex}, where we also use $k=1$ and $k=3$) combined with SSP-RK3 time stepping, and, unless stated otherwise, run all the tests with a CFL factor of $C=0.1$.  
We use the HLL Riemann solver when solving the non-relativistic Euler equations and the HLLC Riemann solver when solving the special relativistic Euler equations.  
The purpose of the tests is to gauge the performance of the DG implementation on a set of benchmarks as an initial measure of its suitability for future CCSN simulations.

@@

The first test is the classic Riemann problem due to Sod \cite{sod_1978}, computed using Cartesian coordinates.  
The one-dimensional computational domain is $D=[0,1]$, and a discontinuity is initially located at $x^{1}=0.5$, separating the left and right states
\begin{equation*}
  \vect{U}_{\leftState}=(1.0,\vect{0.0},2.5)^{\trans}\quad\text{and}\quad
  \vect{U}_{\rightState}=(0.125,\vect{0.0},0.25)^{\trans},
\end{equation*}
where the adiabatic index is $\Gamma=1.4$.  
The test is run with $100$ elements until $t=0.2$, using $C_{\TCI}=0.03$.  
Results are plotted in the left panels of Figure~\ref{fig:SodSedov}.  
The DG method captures the main features of the solution well without introducing noticeable oscillations near the discontinuities (upper left panel).  
In the lower left panel we plot locations in the $xt$-plane of elements flagged for limiting.  
Limiting is mainly triggered by the contact discontinuity and the shock.  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sod_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sedov_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sod_TCI_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Sedov_TCI_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:SodSedov}{\it Upper left panel:} numerical solution of Sod's problem using $100$ elements (plusses) compared with the exact solution (solid lines; from \cite{toro_1999}).  {\it Upper right panel:} numerical solution of the Sedov-Taylor blast wave using $256$ elements (plusses) against the exact solution (solid lines; from \cite{kammTimmes_2007}).  In the lower panels we plot locations in the $xt$-plane of elements flagged for limiting ($I_{\bK}>C_{\TCI}$) by the Sod (left) and Sedov (right) problems.}
\end{figure}

@@

\begin{wrapfigure}{L}{2.825in}
  \includegraphics[width=18pc]{./Figures/IsentropicVortex_Astronum_2018}
  \caption{\label{fig:IsentropicVortex}Convergence results for the Isentropic Vortex:  $L^{\infty}$ error norm versus number of elements in the $x^{1}$-dimension.}
\end{wrapfigure}

In Figure~\ref{fig:IsentropicVortex} we plot the density error measured in the infinity norm $L^{\infty}=\max_{D}|\rho(\vect{x},t_{f})-\rho_{0}(\vect{x})|$ versus number of cells the $x^{1}$-dimension for various configurations.  
We ran with spatial resolutions of $25\times25$, $50\times50$, $100\times100$, and $200\times200$.  
In the first set of runs, we ran without any limiting and varied the polynomial degree $k$: $k=1$ (solid red), $k=2$ (solid black), and $k=3$ (solid green).  
We observe the expected order of accuracy for $k=1$ and $k=2$ (second- and third-order accuracy, respectively; cf. dotted reference lines).  
For $k=3$, the order of accuracy is close to fourth-order for the lower resolutions, but tends to third-order as the resolution increases, presumably due to the use of third-order time stepping.  
(Note that we ran the $k=3$ case with a CFL factor of $1/14$; cf. Eq.~\eqref{eq:timestepRestrictions}.)
In the second set of runs, we kept $k=2$ and ran with limiting turned on.  
We ran one set with $C_{\TCI}=0.03$ (dashed magenta) and one set with $C_{\TCI}=0.1$ (dashed cyan).  
When $C_{\TCI}=0.03$, the limiter is activated for the lowest resolution run and significantly reduces the accuracy of the method.  
For all the other runs, the limiter is not activated, and the results are identical to the runs with the limiter turned off.

@@

\begin{wrapfigure}{R}{2.825in}
  \centering
  \includegraphics[width=18pc]{./Figures/KelvinHelmholtz_TCI_24_Astronum_2018}
  \caption{\label{fig:KelvinHelmholtz_TCI}Distribution of elements flagged for limiting in the KH test at $t=2.4$.}
\end{wrapfigure}

@@

\begin{figure}[b]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Implosion_01_HighTCI_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/Implosion_02_HighTCI_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:Implosion}Plots for the Liska-Wendroff implosion problem at $t=0.045$ (left) and $t=2.5$ (right), intended to match Figures~4.10 and 4.11 in \cite{liskaWendroff_2003}.  The color map shows the pressure distribution, black contours show the mass density, while arrows indicate the velocity.}
\end{figure}
